---
title: "pseudotime"
output: pdf_document
date: "2024-10-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
library(monocle3)
library(ggplot2)
library(dplyr)
# read in data
expr_matrix <- read.csv("./data/no_intensity/ControlData.csv")
expr_matrix2 <- as.matrix(expr_matrix[,-1])
rownames(expr_matrix2) <- as.matrix(expr_matrix[,1]) #setting first column as row names
sample_sheet <- read.csv("./data/no_intensity/Phenodata.csv")
sample_sheet2 <- as.matrix(sample_sheet[,-1])
rownames(sample_sheet2) <- as.matrix(sample_sheet[,1]) #setting first column as row names
gene_annotation <- read.csv("./data/no_intensity/featuredata.csv")
gene_annotation2 <- as.matrix(gene_annotation[,-1])
rownames(gene_annotation2) <- as.matrix(gene_annotation[,1]) #setting first column as row names
cds <- new_cell_data_set(as.matrix(expr_matrix2), cell_metadata = sample_sheet2, gene_metadata = gene_annotation2)
# partition the cds by class
cds_h9 <- cds[, colData(cds)$Class == "H9"]
cds_ucsd <- cds[, colData(cds)$Class != "H9"]


pseudotime_analysis <- function(cds, dim, n_neigh) {
  cds <- preprocess_cds(cds, num_dim = dim)
  cds <- reduce_dimension(cds, preprocess_method = c("PCA"), reduction_method = c("UMAP"))
  cds = cluster_cells(cds, reduction_method = c("UMAP"), k = n_neigh, cluster_method = c("leiden"), resolution=0.1)
  cds <- learn_graph(cds, use_partition = TRUE, close_loop = FALSE, learn_graph_control = NULL, verbose = FALSE)
 # cds <- order_cells(cds)
  return(cds)
}

plot_umap <- function(cds, color_by, discrete_scale = FALSE, legend="Days") {
  plot = plot_cells(cds, reduction_method="UMAP", label_groups_by_cluster = FALSE, label_cell_groups=FALSE,label_leaves=FALSE,
             label_branch_points=FALSE, color_cells_by=color_by) + guides(color = guide_legend(title = legend, override.aes = list(size = 6)))+
    theme(legend.position = "right", legend.key.size = unit(1, 'cm'), legend.text = element_text(size=12,face = "bold"), legend.title = element_text(size=14,face = "bold")) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle(paste("UMAP colored by", legend, ": UCSD Lines (102, 106)"))
  if (discrete_scale) {
    plot = plot + scale_color_viridis_d(option = "plasma")
  }
  return(plot)
}

plot_time_pseudo <- function(cds) {
  plot_data <- data.frame(
  Pseudotime = pseudotime(cds),
  Timepoint = colData(cds)$Dayday,
  Class = colData(cds)$Class
  )

# Convert 'Timepoint' to a factor and then to numeric to calculate correlation
  plot_data <- plot_data %>%
    mutate(Timepoint_numeric = as.numeric(factor(Timepoint, levels = unique(Timepoint))))

# Calculate correlation between Pseudotime and Timepoint for each Class
  stats <- plot_data %>%
    group_by(Class) %>%
    summarize(
      correlation = round(cor(Pseudotime, Timepoint_numeric), 2),
      slope = round(coef(lm(Pseudotime ~ Timepoint_numeric))[2], 2),
      p_value = round(summary(lm(Pseudotime ~ Timepoint_numeric))$coefficients[2, 4], 3)
    )

# Plot box plot with jittered dots and correlation annotation
  ggplot(plot_data, aes(x = Pseudotime, y = Timepoint)) +
    geom_boxplot(aes(fill = Timepoint), outlier.shape = NA) +  # Boxplot without outliers
    geom_jitter(width = 0.2, size = 0.5, alpha = 0.5) +   
    geom_smooth(aes(x = Pseudotime, y = Timepoint_numeric), method = "lm", se = FALSE, color = "black") +# Add jittered points
    theme_minimal() +
    labs(x = "Pseudotime", y = "Timepoint") +
    theme(panel.grid.major = element_line(color = "grey80"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3") +
    facet_wrap(~ Class, scales = "free_y") +                   # Facet by Class
    geom_text(data = stats, aes(x = Inf, y = Inf,
                                label = paste("Corr:", correlation, "\nSlope:", slope, "\nP value:", p_value)),
              hjust = 1.1, vjust = 1.1, size = 4, color = "red") 

}

```

# Data

-   I performed the same analysis for data that exclude *redox* and *intensity* variables and data that only exclude *intensity* variables.

# Pseudotime analysis that excludes intensity and redox vars

-   Since H9 and UCSD lines are quite different, I performed the Pseudo-time analysis separately. 

## Pseudotime analysis on UCSD lines 

First, the two UCSD lines. I tweaked the parameters so that the clustering algorithm does not over-cluster, so that the resulted branches are less. The first output below is the umap plot colored by the day, on top of which is the pseudo-time trajectory. Based on this, I chose the starting point, which is at the right top corner of the umap, the only place I found that have clustered cells from early days (day = 50). 

```{r pseudo_noIR_ucsd}
dim = 8 # only lifetime vars
cds_ucsd <- pseudotime_analysis(cds_ucsd, dim, 500)
plot_umap(cds_ucsd, "Dayday", TRUE)
plot_umap(cds_ucsd, "Class", TRUE, legend="Cell Line")
```

- After choosing the starting cell, it can order the remaining cells based on their distance to the starting cell to get their pseudotime. The second plot below is the umap plot colored by the pseudo-time.
```{r pseudo_noIR_ucsd_p2}
cds_ucsd <- readRDS("cds_ucsd_ordered_noIR.rds")
plot_umap(cds_ucsd, "pseudotime", legend="pseudotime")
```

## Pseudotime analysis on H9 line (no intensity and redox vars)

```{r pseudo_noIR_h9}
cds_h9 <- pseudotime_analysis(cds_h9, dim, 200)
plot_umap(cds_h9, "Dayday", TRUE)
cds_h9 <- readRDS("cds_h9_ordered_noIR.rds")
plot_umap(cds_h9, "pseudotime", legend="pseudotime")
```

## Correlation between pseudo-time and actual time (no intensity and redox vars)
As Andy suggested, we may want to see if the pseudo-time is correlated with the actual time. The plot below shows the correlation between the pseudo-time and the actual time for each class. We can see the similarity of the H9 line and the UCSD 102 line. 
```{r time_vs_pseudotime_noIR}
plot_time_pseudo(cds_ucsd)
plot_time_pseudo(cds_h9)
```

# Pseudotime analysis that excludes intensity vars
``` {r read_data, include=FALSE}
expr_matrix <- read.csv("./data/with_redox/ControlData.csv")
expr_matrix2 <- as.matrix(expr_matrix[,-1])
rownames(expr_matrix2) <- as.matrix(expr_matrix[,1]) #setting first column as row names
sample_sheet <- read.csv("./data/with_redox/Phenodata.csv")
sample_sheet2 <- as.matrix(sample_sheet[,-1])
rownames(sample_sheet2) <- as.matrix(sample_sheet[,1]) #setting first column as row names
gene_annotation <- read.csv("./data/with_redox/featuredata.csv")
gene_annotation2 <- as.matrix(gene_annotation[,-1])
rownames(gene_annotation2) <- as.matrix(gene_annotation[,1]) #setting first column as row names
cds <- new_cell_data_set(as.matrix(expr_matrix2), cell_metadata = sample_sheet2, gene_metadata = gene_annotation2)
# partition the cds by class
cds_h9 <- cds[, colData(cds)$Class == "H9"]
cds_ucsd <- cds[, colData(cds)$Class != "H9"]
cds_102_h9 <- cds[, colData(cds)$Class %in% c("H9", "UCSD102")]
```
- Next, I repeated the same things using the data that include redox. Here are the same set of results. 

## Pseudotime analysis on UCSD lines (no intensity vars)
```{r pseudo_noI}
dim = 9 # lifetime vars and redox
cds_ucsd <- pseudotime_analysis(cds_ucsd, dim, 500)
plot_umap(cds_ucsd, "Dayday", TRUE)
plot_umap(cds_ucsd, "Class", TRUE, legend="Cell Line")
colData(cds_ucsd)$dayClass <- paste(colData(cds_ucsd)$Class,colData(cds_ucsd)$Dayday, sep = "_")
plot_umap(cds_ucsd, "dayClass", TRUE, legend="Class - Day")
cds_ucsd <- readRDS("cds_ucsd_ordered_noI.rds")
plot_umap(cds_ucsd, "pseudotime", legend="pseudotime")
```

## Pseudotime analysis on H9 line (no intensity vars)
```{r pseudo_noI_h9}
cds_h9 <- pseudotime_analysis(cds_h9, dim, 200)
plot_umap(cds_h9, "Dayday", TRUE)
cds_h9 <- readRDS("cds_h9_ordered_noI.rds")
plot_umap(cds_h9, "pseudotime", legend="pseudotime")
```

## Correlation between pseudo-time and actual time (no intensity vars)
```{r time_vs_pseudotime_noI}
plot_time_pseudo(cds_ucsd)
plot_time_pseudo(cds_h9)
```

- My impression is that the pseudo-time here better mimics the actual time compared to the previous analysis. However, the similarity of the correlation between H9 and UCSD 102 lines is not quite there. 

## Pseudotime analysis on UCSD 102 and H9 (no intensity vars)
``` {r pseudo_noI_102_h9}
cds_102_h9 <- pseudotime_analysis(cds_102_h9, dim, 200)
plot_umap(cds_102_h9, "Dayday", TRUE)
plot_umap(cds_102_h9, "Class", TRUE)
colData(cds_102_h9)$dayClass <- paste(colData(cds_102_h9)$Class,colData(cds_102_h9)$Dayday, sep = "_")
plot_umap(cds_102_h9, "dayClass", TRUE, legend="Class - Day")


```
# Final graphs
```{r 1}

cds <- preprocess_cds(cds, num_dim = 9)
plot_umap <- function(cds, color_by, custom_scale = FALSE, legend="Days") {
  custom_colors <- c("H9" = "#FFB6C1", "UCSD102" = "#365ABC", "UCSD106" = "#98FB98")
  plot = plot_cells(cds, reduction_method="UMAP", label_groups_by_cluster = FALSE, label_cell_groups=FALSE,label_leaves=FALSE,
             label_branch_points=FALSE, color_cells_by=color_by, cell_size=0.8, alpha = 0.5) + guides(color = guide_legend(title = legend, override.aes = list(size = 6, alpha=1.0)))+
    theme(legend.position = "right", legend.key.size = unit(1, 'cm'), legend.text = element_text(size=12,face = "bold"), legend.title = element_text(size=14,face = "bold")) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle(paste("UMAP colored by", legend))
  # remove the black outline
  #plot = plot + geom_point(aes(color = factor(cds[[color_by]])), size = 2, stroke = 0, alpha = 0.5)
  
  if (!custom_scale) {
    plot = plot + scale_color_viridis_d(option = "plasma")
  } else {
    plot = plot + scale_color_manual(values = custom_colors)
  }
  return(plot)
}

cds <- reduce_dimension(cds, preprocess_method = c("PCA"), reduction_method = c("UMAP"))
umap_by_class <- plot_umap(cds, "Class", TRUE, legend="Cell Line")
ggsave("umap_by_class.svg", plot = umap_by_class, device = "svg", width = 8, height = 6)
```

```{r 2A}
cds_h9 <- preprocess_cds(cds_h9, num_dim = 9)
cds_h9 <- reduce_dimension(cds_h9, preprocess_method = c("PCA"), reduction_method = c("UMAP"))
plot_umap <- function(cds, color_by, custom_scale = FALSE, legend="Days") {
  custom_colors <- c("H9" = "#FFB6C1", "UCSD102" = "#4682B4", "UCSD106" = "#98FB98")
  plot = plot_cells(cds, reduction_method="UMAP", label_groups_by_cluster = FALSE, label_cell_groups=FALSE,label_leaves=FALSE,
             label_branch_points=FALSE, color_cells_by=color_by, cell_size=1.2, alpha = 0.5) + guides(color = guide_legend(title = legend, override.aes = list(size = 6, alpha=1.0)))+
    theme(legend.position = "right", legend.key.size = unit(1, 'cm'), legend.text = element_text(size=12,face = "bold"), legend.title = element_text(size=14,face = "bold")) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle(paste("UMAP colored by", legend, ": H9 Line"))
  #plot = plot + geom_point(aes(color = factor(cds[[color_by]])), size = 2, stroke = 0, alpha = 0.5)
  
  if (!custom_scale) {
    plot = plot + scale_color_viridis_d(option = "plasma")
  } else {
    plot = plot + scale_color_manual(values = custom_colors)
  }
  return(plot)
}
umap_h9_time <- plot_umap(cds_h9, "Dayday", FALSE, legend="Time in Days")
ggsave("umap_h9_time.svg", plot = umap_h9_time, device = "svg", width = 8, height = 6)
```

```{r 2B}
cds_ucsd <- preprocess_cds(cds_ucsd, num_dim = 9)
cds_ucsd <- reduce_dimension(cds_ucsd, preprocess_method = c("PCA"), reduction_method = c("UMAP"))
plot_umap <- function(cds, color_by, custom_scale = FALSE, legend="Days") {
  custom_colors <- c("UCSD102" = "#365ABC", "UCSD106" = "#98FB98")
  plot = plot_cells(cds, reduction_method="UMAP", label_groups_by_cluster = FALSE, label_cell_groups=FALSE,label_leaves=FALSE,
             label_branch_points=FALSE, color_cells_by=color_by, cell_size=1.0, alpha = 0.5) + guides(color = guide_legend(title = legend, override.aes = list(size = 6, alpha=1.0)))+
    theme(legend.position = "right", legend.key.size = unit(1, 'cm'), legend.text = element_text(size=12,face = "bold"), legend.title = element_text(size=14,face = "bold")) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle(paste("UMAP colored by", legend, ": UCSD Lines (102, 106)"))
  #plot = plot + geom_point(aes(color = factor(cds[[color_by]])), size = 2, stroke = 0, alpha = 0.5)
  
  if (!custom_scale) {
    plot = plot + scale_color_viridis_d(option = "plasma")
  } else {
    plot = plot + scale_color_manual(values = custom_colors)
  }
  return(plot)
}
umap_ucsd_time <- plot_umap(cds_ucsd, "Dayday", FALSE, legend="Time in Days")
ggsave("umap_ucsd_time.svg", plot = umap_ucsd_time, device = "svg", width = 8, height = 6)
```
```{r 2C}
umap_ucsd_class <- plot_umap(cds_ucsd, "Class", TRUE, legend="Cell Line")
ggsave("umap_ucsd_class.svg", plot = umap_ucsd_class, device = "svg", width = 8, height = 6)
```
